---
title: "Preparation"
author: "fionarhuang"
date: "2020-04-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages({
    library(TreeSummarizedExperiment)
    library(treeclimbR)
    library(ape)
    library(ggtree)
    library(ggplot2)
    library(cowplot)
    library(edgeR)
    library(ComplexHeatmap)
    library(dplyr)
    library(TreeHeatmap)
    library(ggnewscale)
    library(scales)
})
```

# Load data
Datasets that are loaded here are prepared by Pierre-Luc ([see here](https://github.com/plger/fiona_miRNA))
```{r}
# count data
heart <- readRDS("data/heart.SE.rds")

# tree
treeO <- readRDS("data/origin_tree.rds")

```


Here, we check whether the mapping is correct between the count table and the tree.
```{r}
# all features could be mapped to the tree
tree_lab <- treeO$tip.label
row_name <- rownames(heart)
all(row_name %in% tree_lab)

# leaves without data
empty_leaf <- setdiff(tree_lab, row_name)
```


# Preprocess tree

Leaves that can't be mapped to rows of the count table are removed.
```{r}
heart_tree <- drop.tip(phy = treeO, tip = empty_leaf, 
                       trim.internal = TRUE, 
                       collapse.singles = TRUE)
heart_tree
```


# Data aggregation

Data on the leaf level is first stored in a
`TreeSummarizedExperiment` container, and is further aggregated to generate
counts on internal nodes of the tree.
```{r}
lse <- TreeSummarizedExperiment(assays = assays(heart),
                                rowData = rowData(heart),
                                colData = colData(heart),
                                rowTree = heart_tree)
# data are on the leaf level
table(rowLinks(lse)$isLeaf)
rowData(lse) <- droplevels(rowData(lse))
```

Threre are 10 samples, five from the Sham surgery and the other five from the TAC surgery. 
```{r}
table(heart$surgery)
```

The aggregation has used the sum of counts on the leaf level.
```{r}

all_node <- showNode(tree = heart_tree, only.leaf = FALSE)
tse <- aggValue(x = lse, rowLevel = all_node,
                FUN = function(x) {
                    sum(x, na.rm = TRUE)
                }, message = TRUE)
saveRDS(tse, file = "output/heart_origin_tse.rds")
```