---
title: "Analysis"
author: "fionarhuang"
date: "2020-04-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages({
    library(TreeSummarizedExperiment)
    library(treeclimbR)
    library(ape)
    library(ggtree)
    library(ggplot2)
    library(cowplot)
    library(edgeR)
    library(ComplexHeatmap)
    library(dplyr)
    library(TreeHeatmap)
    library(ggnewscale)
    library(scales)
})
```

# Load data

```{r}
tse <- readRDS("output/heart_origin_tse.rds")
```


# Differential analysis

`edgeR` was used to perform differential abundance (DA) test on all nodes..

```{r}
surgery <- colData(tse)$surgery
surgery <- ifelse(surgery == "Sham", "Sham", "TAC")
names(surgery) <- colnames(tse)

res <- runDA(TSE = tse, feature_on_row = TRUE, assay = 1,
             option = "glm", group_column = "surgery",
             design_terms = "surgery")
out <- nodeResult(object = res, n = Inf)
```


## Run `treeclimbR`

Multiple candidates are proposed by tuning `t`.

```{r}
cand <- getCand(tree = rowTree(tse), 
                score_data = out, node_column = "node",
                p_column = "PValue", sign_column = "logFC", 
                message = FALSE)
```

All proposed candidates are evaluated. One candidate is selected as the best at FDR $\alpha = 0.05$.
```{r}
candB <- evalCand(tree = rowTree(tse), 
                  levels = cand$candidate_list, 
                  score_data = out, node_column = "node", 
                  p_column = "PValue", sign_column = "logFC",
                  method = "BH", limit_rej = 0.05,
                  use_pseudo_leaf = FALSE,
                  message = FALSE)
```


# Results

Nodes that are reported to be DA are in `loc_0.05`.
```{r}
# the result table
result <- topNodes(object = candB, n = Inf, p_value = 0.05)
loc_0.05 <- result$node

# the logFC of reported nodes
loc_lfc <- result$logFC

# the sign of reported nodes
loc_sign <- ifelse(sign(loc_lfc) > 0, "Up-regulated", "Down-regulated")
names(loc_0.05) <- loc_sign

head(loc_0.05)
```


The size (number of leaves) of identified branches.
```{r}
leaf_0.05 <- findOS(tree = rowTree(tse), node = loc_0.05, 
                    only.leaf = TRUE, self.include = TRUE)
size_0.05 <- unlist(lapply(leaf_0.05, length))
table(size_0.05)

# the totoal number of leaves covered by identified nodes
sum(size_0.05)
```


```{r}
# the number of nodes detected:
length(loc_0.05)

# the number of sequences detected
sum(size_0.05)
```


Below gives miRNAs that the identified sequences belong to:
```{r}
seqs <- unlist(leaf_0.05)
seqs_mir <- rowData(tse)[rowLinks(tse)$nodeNum %in% seqs,
                                "Annotation"]
seqs_mir <- unique(as.character(seqs_mir))
head(seqs_mir)
# For those with more than one mirs separated by ";"
# For example, "a;b"means sequences could be from a or b or both
seqs_mir <- unlist(strsplit(seqs_mir, ";"))
seqs_mir <- unique(seqs_mir)

# the number of miRNAs
length(seqs_mir)
head(seqs_mir)
```

```{r}
saveRDS(candB, file = "output/heart_origin_result.rds")
```