---
title: "Visualization"
author: "fionarhuang"
date: "2020-04-24"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
suppressPackageStartupMessages({
  library(TreeSummarizedExperiment)
  library(dplyr)
  library(treeclimbR)
  library(ggtree)
  library(TreeHeatmap)
  library(ggplot2)
  library(viridis)
  library(ggnewscale)
  library(edgeR)
  library(scales)
  library(grid)
  library(cowplot)
  library(png)
})
```


```{r}
tse <- readRDS("output/heart_origin_tse.rds")
candB <- readRDS("output/heart_origin_result.rds")
# information about sequence 
origin_df <- readRDS("data/origin_dataframe.rds")
```


# Plot tree

```{r}
# tree
heart_tree <- rowTree(tse)

# nodes
all_node <- showNode(tree = heart_tree, only.leaf = FALSE)

# figures
fig_tree <- ggtree(heart_tree, ladderize = FALSE,
                   layout = "fan", 
                   open.angle = 10,
                   size = 0.5)
fig_tree
```

# Color branches
```{r}
# prepare data to color branches
result <- topNodes(object = candB, n = Inf, p_value = 0.05)
loc_0.05 <- result$node
names(loc_0.05) <- ifelse(result$logFC > 0, "Up-regulated", "Down-regulated")

br_0.05 <- findOS(tree = heart_tree, node = loc_0.05,
                  only.leaf = FALSE, self.include = TRUE)
df_branch <- data.frame(node = unlist(br_0.05),
                        Branch = rep(names(loc_0.05),
                                     unlist(lapply(br_0.05, length)))) %>%
    distinct() %>%
    add_row(node = setdiff(all_node, unlist(br_0.05)),
            Branch = "same")

# branch colors
branch_col <- c("same" = "grey70", "Up-regulated" = "navy", 
                "Down-regulated" = "orange")

# figure: tree
fig_tree <- fig_tree %<+% 
    df_branch + 
    aes(color = Branch) + 
    scale_color_manual(values = branch_col, 
                       guide = guide_legend(
                         override.aes = list(size = 3),
                         order = 1)) +
    geom_point2(aes(subset = (node %in% loc_0.05)), 
                color = "red", shape = 19, size = 1.5) 

fig_tree

```

# Scale branches

Branches without signal are shrinked to save space.
```{r}
# branches without signal
leaf <- showNode(tree = heart_tree, only.leaf = TRUE)
leaf_yes <- findOS(tree = heart_tree, node = loc_0.05, 
                   only.leaf = TRUE, self.include = TRUE)
leaf_no <- setdiff(leaf, unlist(leaf_yes))
node_no <- signalNode(tree = heart_tree, node = leaf_no)
node_no <- node_no[!isLeaf(tree = heart_tree, node = node_no)]


circo_full <- fig_tree
for (i in node_no) {
    circo_full <- scaleClade(circo_full, node = i, scale = 0.2)
}
circo_full
```

Branches that have signal and have more than 10 leaves are scaled to have more space.
```{r}
# zoom in signal branches
loc_a <- findOS(tree = heart_tree, node = loc_0.05, 
                only.leaf = TRUE, self.include = TRUE)
len <- unlist(lapply(loc_a, length))
node_a <- loc_0.05[len > 10]

for (i in node_a) {
    circo_full <- scaleClade(circo_full, node = i, scale = 3.5)
}

```


# Label hierarchical levels

```{r}
ux <- setdiff(sort(unique(circo_full$data$x)), 0)
names(ux) <- c("genomic_cluster", "primary_tx", "miRNA", "sequence" )
df_layer <- data.frame(x = ux)%>%
    arrange(x) %>%
    mutate(y = max(circo_full$data$y, na.rm = TRUE),
           layer = factor(names(ux), levels = names(ux)))

# annotate the four levels on the tree
layer_shape <- 49:52
names(layer_shape) <- names(ux)
circo_full <- circo_full + 
    geom_point(data = df_layer, 
               aes(x, y = y + 210, shape = layer),
               size = 6, 
               inherit.aes = FALSE, show.legend = FALSE) +
    scale_shape_manual(values = layer_shape, 
                       guide = guide_legend(
                           override.aes = list(size = 4), 
                           order = 1)) +
    labs(shape = "Hierarchy")
circo_full  
```


# Heatmap

The heatmap uses data on the leaf level. Counts of miRNAs in cpm are firstly log transformed, and then standardized using `scale` for each miRNA across samples.

```{r}
# the data on the leaf level for signal branches
yes_L <- findOS(tree = heart_tree, node = loc_0.05, 
                only.leaf = TRUE, self.include = TRUE)
yes_L <- rowLinks(tse)$nodeNum %in% unlist(yes_L)

# scale counts:
#   1. log transformed in cpm counts
#   2. log-count of each miRNA is further scaled among samples

# log-transformed by columns
count_L <- assays(tse)[[1]][yes_L, ]
countL <- cpm(count_L)
scale_L <- log(countL + 1) 

# standardized
scale_L <- t(apply(scale_L, 1, FUN = function(y) {
    ly <- length(unique(y))
    if (ly == 1) {
        sy <- rep(0, length(y))
    } else {
        sy <- scale(y)
    }
    return(sy)
}))
rownames(scale_L) <- rownames(count_L)
colnames(scale_L) <- colnames(count_L)
dim(scale_L)
```

Heatmap is only on leaves that are identified as differentially abundant.
```{r}
surgery <- colData(tse)$surgery
surgery <- ifelse(surgery == "Sham", "Sham", "TAC")
names(surgery) <- colnames(tse)

fig2 <- TreeHeatmap(tree = heart_tree, 
                    rel_width = 0.2,
                    tree_fig = circo_full,
                    hm_data = scale_L, 
                    legend_title_hm = "Expression(scale)",
                    tree_hm_gap = 0.3,
                    column_split = surgery, 
                    column_split_gap = 0.05, 
                    column_split_label = c("Sham" = "Sham", 
                                           "TAC" = "TAC"), 
                    split_label_angle = 80,
                    split_label_size = 4, 
                    split_label_offset_y = 260) +
    scale_fill_gradient(low = "navy", high = "yellow",
                        limits = c(-1, 1),
                        oob = squish,
                        guide = guide_colorbar(order = 3)) +
    new_scale_fill() 

fig2
```

# Annotation: miRNA

```{r}
leaf_a  <- findOS(tree = heart_tree, node = node_a, 
                  only.leaf = TRUE, self.include = TRUE)

names(leaf_a) <- transNode(tree = heart_tree, node = node_a,
                           use.alias = TRUE)

df_a <- lapply(node_a, FUN = function(x){
    xx <- rowData(tse)[rowLinks(tse)$nodeNum %in% x, ]
})


mir_a <- lapply(leaf_a, FUN = function(x){
    ix <- (rowLinks(tse)$nodeNum %in% x) & rowLinks(tse)$isLeaf
    xx <- rowData(tse)[ix, "Annotation"]
    unique(as.character(xx))
})
head(mir_a)


mir_s_a <- lapply(mir_a, FUN = function(x){
    
    if (length(x) == 1) {
        gsub(pattern = "mmu-mir-", "", x)
    } else {
        xx <- unique(stringr::str_extract(x, "^mmu-mir-[0-9]+"))
        if (length(xx) == 1) {
            x <- unlist(strsplit(x = x, split = ";"))
            x <- unique(x)
            px <- paste(x, collapse = ";")
            gsub(pattern = "mmu-mir-", "", px)
        } else {
            "mixed" 
        }}
})

mat_a <- data.frame(mir = (unlist(mir_s_a)))
rownames(mat_a) <- names(mir_s_a)

length(unique(mat_a[, 1]))


```


```{r, fig.height=6.5, fig.width=7.8}
dat <- getData(tree_hm = fig2, type = "heatmap")
df_line <- data.frame(x = max(dat$x)+0.25,
                      y1 = max(min(dat$y), 0), 
                      y2 = max(dat$y))

cols <- c("darkmagenta", "darkorchid1", "cornflowerblue", "cyan3", 
          "blue", "plum2", "orange", "firebrick2", "tan4",
          "deeppink", "darkslateblue", "darkred", "magenta",
          "darkseagreen",
          "mediumpurple", "darkolivegreen1", "gray", "black")
names(cols) <- sort(unique(mat_a[, 1]))


fig3 <- TreeHeatmap(tree = heart_tree, 
                    tree_fig = fig2,  
                    tree_hm_gap = 1.5,
                    rel_width = 0.05, 
                    hm_data = mat_a,
                    legend_title_hm = "miRNA") +
    scale_fill_manual(values = cols, 
                      guide = guide_legend(order = 4, ncol = 2)) +
    geom_segment(data = df_line,
                 aes(x = x, y = y1, xend = x, yend = y2),
                 linetype = "dashed",  
                 inherit.aes = FALSE)  

fig3 +
  theme(
        aspect.ratio = 1,
        legend.position = c(0.12, 0.5),
        legend.background = element_rect(fill = NA),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 9),
        legend.text = element_text(size = 7),
        legend.key.size = unit(0.3, "cm"),
        legend.spacing.x = unit(0.2, "mm"),
        legend.margin = margin(t = 0, b = 0, r = 20, l = 0),
        legend.box.margin=margin(t = -5, b = -5, r = 25, l = -60),
        plot.margin = margin(t = -50, b = -50, r = -150, l = -40)
    )

```

# Annotate genomic clusters

Identified nodes are annotated with their correspoinding hierarchy levels.
```{r}
source("code/fun_inner_label.R")
# leaf labels
leaf_0.05 <- findOS(tree = heart_tree, node = loc_0.05, 
                    only.leaf = TRUE, self.include = TRUE)
label_0.05 <- transNode(tree = heart_tree, node = unlist(leaf_0.05))

# hierarchical levels of detected nodes
node_info <- origin_df %>% 
    filter(sequence %in% label_0.05)
node_lab <- fun_inner_label(tree = heart_tree, 
                            tree_df = node_info, 
                            column_leaf = "sequence")
lab_0.05 <- lapply(node_lab, FUN = function(x) {
    x[x %in% loc_0.05]
})

# the number of nodes detected at each level
lapply(lab_0.05, length)

# on genomic_cluster level
lab_0.05$`1`
```



```{r, fig.height=6.5, fig.width=7.8}
# data to plot points
dat <- getData(tree_hm = fig3, type = "heatmap")

df_point <- data.frame(node = lab_0.05$`1`, 
                    "Genomic_range" = factor(names(lab_0.05$`1`)))  %>%
  left_join(dat) %>%
  mutate(x = x + 0.2, y = y )

# colors for points
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
col_g <- cols[df_point$value]
col_g[col_g == "black"] <- cbPalette[seq_len(sum(col_g == "black"))]
names(col_g) <- df_point$Genomic_range

fig4 <- fig3 +
  new_scale_color() +
  geom_point(data = df_point, 
             aes(x = x, y = y, color = Genomic_range),
             shape = 19, size = 4,
            inherit.aes = FALSE) +
  scale_color_manual(values = col_g) +
  guides(color = guide_legend(order = 5, ncol = 2))+
    theme(
        aspect.ratio = 1,
        legend.position = c(0.20, 0.33),
        legend.background = element_rect(fill = NA),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 6),
        legend.key.size = unit(0.25, "cm"),
        legend.spacing.x = unit(0.2, "mm"),
        legend.margin = margin(t = 0, b = 0, r = 50, l = 0),
        legend.box.margin=margin(t = -5, b = -5, r = 25, l = -60),
        plot.margin = margin(t = -50, b = -50, r = -150, l = -40)
    )
fig4 
ggsave("output/heart_microRNA.eps", fig4, 
       units = "in", width = 7.8, height = 6.5,
       dpi = 300)

```

# Add hierarchy in legend
```{r, fig.height=6.5, fig.width=7.8}
img <- readPNG("data/hierarchy.png")
g <- rasterGrob(img, width = 4.265, height = 4.453,
                default.units = "cm",
                interpolate=FALSE)


figF <- ggdraw(fig4) +
    draw_grob(g, x = -0.38, y = 0.3, scale = 1)
figF
ggsave("output/heart_microRNA.pdf", figF,
       units = "in", width = 7.8, height = 6.5,
       dpi = 300)
```